Theory of phonon polaritons in polar dielectrics

### 2018-08-03 paper.tex
minor tweak... now i'm done :)

\begin{abstract}
The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.  An important recent development along this paradigm is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band. To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles. In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function. For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model. However, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  and the phonon-bandstructure.  Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm. Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice. We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion. Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide and hexagonal boron nitride, and show that they in general agree well with experimental data. Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
\end{abstract}

@@

The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.

@@

To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density. We have that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
\end{equation}
where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons. Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations. We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged. And instead of the solving Maxwell's equations for the field, one solves Laplace's equation for the potential.  Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
\end{equation}
leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.

@@

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra. % For a material in which the band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used for an entirely \it{ab initio} implementation of our theory. If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy. 
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good.

>>>>

\begin{abstract}
The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.  An important recent development along this paradigm is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band. To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles. In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function. For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model. However, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  and the phonon-bandstructure.  Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm. Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice.  To show the practical usage of the framework derived, we provide ab-initio calculations for bulk silicon carbide and bulk hexagonal boron nitride, and show that they in general agree well with the measured values of their dielectric functions. Finally, to show the generality of the framework developed here, we derive the electrodynamics of 2D phonon polaritons, finding a universal form for their electromagnetic modes and dispersion relation, which can be specified in terms of ab-initio parameters. Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
\end{abstract}

@@

The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant. We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements. Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of ab-initio parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.

@@

To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density. We have that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
\end{equation}
where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons. Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations. This is essentially the random-phase approximation (RPA). We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged. And instead of the solving Maxwell's equations for the field, one solves Laplace's equation for the potential.  Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
\end{equation}
leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.

@@

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra. % For a material in which the band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used for an entirely \it{ab initio} implementation of our theory. If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy. 
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN. For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.
